---
title: "Case 1: The Lucahura Sales Drop"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case 1: The Lucahura Sales Drop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Prologue: the case

You are a data analyst at **Lucahura Retail Group**, a retail chain with
9 stores spread across three regions: *North*, *Central*, and *South*.

The regional manager writes to you, worried: sales in one of the three
regions dropped sharply in the third quarter of 2025, and the
meeting room is already full of theories about why:

1. "We raised prices and people stopped buying."
2. "We cut the marketing budget."
3. "We lost some of our key salespeople."
4. "A competitor opened nearby."

Your job is to use R to check the data, rule out the theories that
don't hold up, and keep the explanation the numbers actually support.
You'll need six tables, all included in this package: `sales`,
`stores`, `products`, `employees`, `marketing`, and `competitors`.

Before testing a single hypothesis, load the package and take a first
look at the data with `glimpse()`:

```{r setup}
library(datamysteries)
library(dplyr)
library(ggplot2)
library(lubridate)
library(stringr)

dplyr::glimpse(sales)
dplyr::glimpse(stores)
```

In an interactive RStudio session you can also run `View(sales)` to
explore the full table in a separate pane.

**Reflection question:** before writing a single line of analysis,
which of the four theories seems most likely to you, just on
intuition? Write it down — at the end of the case you'll be able to
compare your hunch against the evidence.

---

## Chapter 1: where is the drop?

Before checking causes, first confirm whether the drop exists, when it
starts, and which region it is concentrated in. To do that, combine
`sales` with `stores`, summarize sales by region and month, and compare
the months around the third quarter of 2025.

```{r cap1-by-region}
sales_by_region <- sales |>
  dplyr::left_join(stores, by = "store_id") |>
  dplyr::mutate(year_month = format(date, "%Y-%m")) |>
  dplyr::group_by(region, year_month) |>
  dplyr::summarise(total = sum(amount), .groups = "drop")

sales_by_region |>
  dplyr::filter(year_month >= "2025-05") |>
  tidyr::pivot_wider(names_from = region, values_from = total)
```

Now calculate the percentage change from June to July 2025 for each
region. The region with the most negative change is your main suspect.

```{r cap1-affected-region}
affected_region <- sales_by_region |>
  dplyr::filter(year_month %in% c("2025-06", "2025-07")) |>
  dplyr::arrange(region, year_month) |>
  dplyr::mutate(
    pct_change = (total / dplyr::lag(total) - 1) * 100,
    .by = region
  ) |>
  dplyr::filter(year_month == "2025-07") |>
  dplyr::arrange(pct_change) |>
  dplyr::slice(1) |>
  dplyr::pull(region)

affected_region
```

**Finding:** write down, in your own words, which region shows the
break and how different its behavior is from the other regions.

**Reflection question:** before looking for causes, what minimum
evidence do you need to say this is a regional problem rather than a
company-wide problem?

---

## Chapter 2: did prices go up?

The most repeated theory in the meeting room is about prices. You have
now identified the affected region; next, check whether the price
information can explain a localized drop.

```{r cap2-prices}
products |>
  dplyr::select(product_id, category, list_price)
```

```{r cap2-affected-region-stores}
affected_region_stores <- stores |>
  dplyr::filter(region == affected_region)

affected_region_stores
```

**Finding:** the `products` table has a single list price per product
for the *whole* company; there is no region or store column. In other
words, if prices went up, they went up equally in all 9 stores, not
just in the affected region.

**Reflection question:** if a price change applies equally to every
region, can it, by itself, explain a sales drop that only happens in
one region?

---

## Chapter 3: quantifying the change with `mutate()` and `arrange()`

The `products` table only has the *current* price. To find out whether
the price really went up, and when, we need to calculate it from
`sales`, dividing the amount sold by the units: `amount / quantity`.

```{r cap3-unit-price}
unit_price_by_month <- sales |>
  dplyr::left_join(stores, by = "store_id") |>
  dplyr::mutate(
    unit_price = amount / quantity,
    year_month = format(date, "%Y-%m")
  ) |>
  dplyr::group_by(region, year_month) |>
  dplyr::summarise(avg_price = mean(unit_price), .groups = "drop")

unit_price_by_month |>
  dplyr::arrange(year_month) |>
  head(12)
```

Now let's calculate the month-over-month percentage change in sales by
store, and sort from biggest to smallest drop:

```{r cap3-pct-change}
sales |>
  dplyr::left_join(stores, by = "store_id") |>
  dplyr::group_by(store_id, region, date) |>
  dplyr::summarise(monthly_amount = sum(amount), .groups = "drop") |>
  dplyr::arrange(store_id, date) |>
  dplyr::mutate(
    pct_change = (monthly_amount / dplyr::lag(monthly_amount) - 1) * 100,
    .by = store_id
  ) |>
  dplyr::filter(!is.na(pct_change)) |>
  dplyr::arrange(pct_change) |>
  head(10)
```

**Finding:** the average unit price rises at some point in 2025 — but
comparing `region` against `year_month`, the increase happens **in all
three regions at the same time**. Meanwhile, the sharpest percentage
drops in `pct_change` are concentrated in stores from a single region.

**Reflection question:** with this evidence, is the price theory ruled
out, confirmed, or does it need more data?

---

## Chapter 4: comparing regions with `group_by()` and `summarize()`

Let's confirm the pattern above more directly: total sales by region and
month.

```{r cap4-by-region}
sales_by_region <- sales |>
  dplyr::left_join(stores, by = "store_id") |>
  dplyr::mutate(year_month = format(date, "%Y-%m")) |>
  dplyr::group_by(region, year_month) |>
  dplyr::summarise(total = sum(amount), .groups = "drop")

sales_by_region |>
  dplyr::filter(year_month >= "2025-05") |>
  tidyr::pivot_wider(names_from = region, values_from = total)
```

**Finding:** write down, in your own words, in which month and in which
region the break happens, and how the other two regions behave over the
same period.

**Reflection question:** with the price theory weakened, which three
hypotheses are left to check? Revisit the list from the prologue.

---

## Chapter 5: joining in staffing data (`left_join()`)

One of the theories from the meeting room is about the departure of key
salespeople. Let's join `employees` with `stores` to look at turnover by
region.

```{r cap5-employees}
departures_2025 <- employees |>
  dplyr::left_join(stores, by = "store_id") |>
  dplyr::filter(!is.na(termination_date), termination_date >= as.Date("2025-01-01"))

departures_2025 |>
  dplyr::count(region, sort = TRUE)
```

```{r cap5-dates}
departures_2025 |>
  dplyr::filter(region == "South") |>
  dplyr::arrange(termination_date) |>
  dplyr::select(employee_id, store_id, hire_date, termination_date)
```

**Finding:** note how many departures happened in each region during
2025, and in which months they're concentrated for the most affected
region.

**Reflection question:** do the departure dates happen before, during,
or after the sales drop you identified in Chapter 4?

---

## Chapter 6: budget and competition (`str_detect()` and dates)

Two hypotheses remain: the marketing cut and the competitor opening.
Let's start with marketing, comparing the suspect region's budget
against the others:

```{r cap6-marketing}
marketing |>
  dplyr::filter(start_date >= as.Date("2025-01-01")) |>
  dplyr::arrange(region, start_date) |>
  dplyr::select(region, start_date, budget)
```

Now let's look at `competitors`. Use `str_detect()` on the competitor's
name if you want to filter by keyword, and `lubridate` functions to
compare how recent each opening is:

```{r cap6-competition}
competitors |>
  dplyr::mutate(
    years_since_opening = lubridate::interval(opening_date, as.Date("2025-12-01")) %/% months(12)
  ) |>
  dplyr::arrange(opening_date)
```

```{r cap6-strdetect}
competitors |>
  dplyr::filter(stringr::str_detect(competitor, "Max|Value|Quick"))
```

**Finding:** compare the affected region's marketing budget against its
own history (did it drop compared to previous quarters?) and against the
other regions. Then compare how old each competitor is: which one
opened years ago, and which one opened just a few months ago?

**Reflection question:** of the four original theories, which can you
rule out with the evidence gathered so far, and which are still
standing?

---

## Chapter 7: visualizing the finding with `ggplot2`

A chart tends to convince a manager faster than a table. Plot the
monthly sales trend by region:

```{r cap7-plot, fig.width=7, fig.height=4}
sales |>
  dplyr::left_join(stores, by = "store_id") |>
  dplyr::group_by(region, date) |>
  dplyr::summarise(total = sum(amount), .groups = "drop") |>
  ggplot2::ggplot(ggplot2::aes(x = date, y = total, color = region)) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::labs(
    title = "Monthly sales by region",
    x = NULL, y = "Sales amount", color = "Region"
  ) +
  ggplot2::theme_minimal()
```

**Finding:** describe in your own words what the chart shows. In which
month does the affected region's line separate from the other two?

---

## Final chapter: putting the whole case together

You now have evidence from five different angles: prices, sales by
region, staff turnover, marketing budget, and competitor openings. Now
it's time to combine them into a single summary table for the
struggling region, and use it to support your conclusion.

```{r cap-final-summary}
south_summary <- sales |>
  dplyr::left_join(stores, by = "store_id") |>
  dplyr::filter(region == "South") |>
  dplyr::group_by(date) |>
  dplyr::summarise(monthly_sales = sum(amount), .groups = "drop") |>
  dplyr::mutate(period = dplyr::if_else(date < as.Date("2025-07-01"), "before", "after")) |>
  dplyr::group_by(period) |>
  dplyr::summarise(avg_monthly_sales = mean(monthly_sales), .groups = "drop")

south_summary
```

With that table, plus what you already found in `employees` and
`competitors` for the South region, you have everything you need to
answer as a business analyst:

**Executive decision:** the regional manager has budget for one
intervention over the next 60 days. Write a recommendation in two or
three sentences: what you believe is the real cause of the drop, which
theories you ruled out and why, which action you would prioritize, and
which metric you would use to evaluate whether the intervention worked.

> If you want to compare your conclusion against an instructor's, the
> package repository includes a `solutions/case1_solution.R` file (not
> installed with the package, available on GitHub) with the full
> analysis and the final reveal — avoid it until you've written your
> own answer.
